Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-10 intro.tex
adding Liz's edits to intro

Complicated time series are ubiquitous in modern scientific research.
Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}. On the low end of that spectrum are time series that exhibit perfect predictive structure, i.e, signals whose future value are perfectly predicted by knowledge of past values from some ``ideal" model. In particular, there is a an underlying process that generates and transmits information from the past to the future in a perfectly predictable fashion. Some examples of this are constant or low-period signals. On the opposite end of this spectrum are signals which are ``fully complex", i.e., the underlying generating process creates information so rapidly that no information at all is transmitted from the past to the future. Some examples, of this are white noise or random walk processes. With signals on this side of the spectrum, knowledge of the past gives absolutely no insight into the future, \emph{regardless} of the chosen model.  In the middle of this spectrum are where complicated \emph{and} interesting---from a modeling perspective---signals exist, e.g., observations of a deterministic chaotic trajectory, high period signals with some level of noise on top, etc. With time series in this portion of the spectrum, enough information is being transmitted from the past to the future that given the \emph{ideal} model the future of the observed system could be forecast with high accuracy.

Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{weigend-book}, let alone an arbitrary time series. This naturally leads to an important and hard question: given a complicated real-valued time series, does their exist a forecast model for that time series that can utilize the information being transmitted from past to future (if such information exists)? A first step in answering this question is to answer: is it possible to reliably quantify where on the spectrum of complexity the time series lies? i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast? Since the information in an observation can be partitioned into two pieces --- redundancy and entropy generation\cite{crutchfield2003} --- our goal is to analyze how the amount of predictive structure present in a signal, as measured by the redundancy, relates to the performance of prediction strategies, as well as what is meant by \emph{low, medium and high} complexity in Section \ref{sec:meaComplex}.

To answer this question and for the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which appear \emph{complicated}, i.e., they exhibit interesting behavior, for instance, not of low period, strictly monotonic, or constant. We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
We argue that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy of a real-valued-finite-length-noisy time series through ordinal analysis---is an effective way to explore that conjecture. In light of this, for the purposes of this paper we define \emph{complexity} as an approximation of Shannon \emph{entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}. In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity. The permutation entropy is ideal in a situation like this as it converges on the true entropy value, unlike other partitioning techniques used in the past which either require specific knowledge of the generating process or produce biased values of the entropy~\cite{bollt2001}.

@@

To explore this conjecture we require a test bed for generating a broad array of time series which cover the full spectrum of complexity. We argue that the ideal test bed for this purpose are computer performance traces, e.g., the processor efficiency during the execution of a particular program. Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor efficiency highly complicated.
%Accurate forecasts of these quantities, if one could
%construct them, could be used to improve computer design.
% If one
%could predict that a particular computational thread %would be bogged
%down for the next 0.6 seconds waiting for data from main memory, for
%instance, one could save power by putting that thread %on hold for that
%time period (e.g., by migrating it to a processing %unit whose clock
%speed is scaled back).

As a result computer performance traces are very
complicated and range across the entire spectrum of complexity, making this an ideal test bed.  Even a simple ``microkernel," like a three-line loop that
repeatedly initializes the upper triangle of a matrix in column-major order, can produce everything from periodic to
chaotic performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
Figure~\ref{fig:ipc}.%, and chaos places fundamental limits on predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of \col, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
More interesting programs, e.g., compilers and numerical solvers, produce time series which exist throughout the spectrum of complexity: from completely predictable to completely unstructured,i.e., low to high complexity[[Just trying to reinforce the wording in the readers mind]]. In addition, some performance traces even exhibit interesting regime changes as a program switches between varying subroutines. With a single system (computer performance) capable of producing complicated time series that vary so greatly in complexity (periodic to noise), it would be invaluable to known a priori if a trace contained enough predictive structure to build a  forecast model based on structure before time and effort was spent searching for an ideal model.

@@

This paper provides insight into when, why, and how forecast strategies fail when they are applied to complicated time series by empirically estimating their inherent complexity. [[Liz, do you think here would be better for the sentence about we are not the first, or is above where I asked Ryan to insert it ok? I worry it would muddy the water of this synopsis paragraph.]] We focus here on the specific example of
computer performance but believe the results apply to a much broader class of time series. We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity, as well as from
measurement issues like noise, aggregation, and finite data
length---is \emph{empirically quantifiable}. To validate this conjecture we model 120 different performance traces using a variety of forecasting models which we introduce in Section \ref{sec:model}. We then compare the empirically estimated \emph{complexity} with the accuracy of each forecast method. Which results in two primary findings:

\begin{enumerate}

\item The complexity of a noisy real-valued time series is quantifiable and correlated with prediction accuracy of an ideal predictor.

\item The way information is processed internally by a given process plays a crucial role in the success of different forecasting schema.

\end{enumerate}

@@

% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU, e.g., processor efficiency and similar statistics. Prediction of these quantities
at some finite time in the future, even with perfect accuracy, does
not violate the Rice-Shapiro theorem~\cite{hopcroft} which essentially states that any non-trivial property of a program is uncomputable.

 The rest of the paper is organized as follows: Section~\ref{sec:related} discusses similar methods and how this method is different. Section~\ref{sec:methods} explains the experimental setup and explains in depth the programs observed to generate each time series. Section~\ref{sec:model} describes the four forecast models we implemented.  In
 Section~\ref{sec:meaComplex}, we review weighted permutation entropy. In Section~\ref{sec:results} we estimate the complexity of each time series, and
 compare the complexity to the prediction accuracy of several different forecast schema.  In
 Section~\ref{sec:conc}, we discuss these results and their
 implications in regard to our conjecture, and consider future areas of
 research.

>>>>

Complicated time-series data are ubiquitous in modern scientific
research.  The complexity of these data spans a wide range.  On the
low end of this spectrum are time series that exhibit perfect
predictive structure, i.e, signals whose future values can be
successfully predicted from past values.  In situations like this,
there is an underlying process that generates information and/or
transmits it from the past to the future in a perfectly predictable
fashion.  Constant or periodic signals, for example, fall in this
class.  On the opposite end of this spectrum are signals that are what
we call \emph{fully complex}, where the underlying generating process
transmits no information at all from the past to the future.  Some
examples of this are white noise or random-walk processes.  In signals
like this, knowledge of the past gives no insight into the future,
\emph{regardless} of what model one chooses to use.  Signals in the
midrange of this spectrum---e.g., deterministic chaos, high-period
signals [[Liz question: why does the rank of the periodicity force
    them off the low end of the spectrum?  Any periodic signal can be
    decomposed into sine waves, which can be predicted perfectly...]]
with some additive noise---pose interesting challenges from a modeling
perspective.  In these signals, enough information is being
transmitted from the past to the future that an \emph{ideal}
model---one that captures the generating process---can forecast the
future behavior of the observed system with high accuracy.  [[Liz
    question: if there's noise, is that true?]]

Unfortunately, as a corollary of the undecidability of the halting
problem \cite{halting-problem}, there cannot exist \emph{one}
[[universally]] [[across {\sl all} time series?  this seems pretty
    obvious...]] ideal forecasting scheme for even completely
noise-free deterministic time series\cite{weigend-book}---let alone an
arbitrary real-world time series.  This naturally leads to an
important and challenging question: given a complicated real-valued
time series, does there exist a forecast model that can utilize the
information (if any) that is being transmitted from past to future by
the underlying generating process?  A first step in answering this
question is to reliably quantify where on the complexity spectrum a
given time series falls; a second step is to determine how complexity
and predictability are related.  These are the goals of this paper.

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to
predict.  An important practical corollary to this is assessment of
forecast methods: if the forecast produced by a particular method is
poor but the time series contains a significant amount of predictive
structure, the method is inadequate and one should seek another
method.

We focus on real-valued, scalar, time-series data that appear
\emph{complicated}: i.e., not periodic, linear, constant, etc.  We
make no assumptions about the properties of the generating process:
whether it is linear, nonlinear, deterministic, stochastic,
stationary, non-stationary, etc.  The information in an observation
can be partitioned into two pieces: redundancy and entropy
generation\cite{crutchfield2003}.  Our approach exploits this
decomposition in order to assess how much predictive structure is
present in a signal---i.e., where it falls on the complexity spectrum
mentioned above.  We define \emph{complexity} as a particular
approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.  That is,
we view a random-walk time series (which exhibits high entropy) as
purely complex, whereas a periodic signal, which exhibits low entropy,
is on the low end of the complexity spectrum.  We argue that an
extension of \emph{permutation entropy} \cite{bandt2002per}---a method
for approximating the entropy through ordinal analysis---is an
effective way to assess the complexity of a given time series.
Permutation entropy is ideal for our purposes because it is known to
converge to the true entropy value, unlike other existing techniques,
which either require specific knowledge of the generating process or
produce biased values of the entropy~\cite{bollt2001}.

To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a number of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is
  quantifiable by permutation entropy and correlated with prediction
  accuracy of an ideal predictor [[we don't know that.  we only tried
      four methods.]]
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema.
\end{enumerate}
The forecast methods used here were chosen to span the space of
standard prediction strategies, but they do not cover that space
exhaustively.  Our goal here is an empirical assessment of the
relationship between predictability and complexity, not formal results
about a ``best'' predictor for a given time series.  From a
practitioner's standpoint, it would be useful to know {\sl a priori}
if a trace contains enough predictive structure to make it worth
spending the time and effort searching for a good forecast method.  It
would also be useful to know if a given method is inadequate---that
is, if one could do better and should keep searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  Computers are among the most complex
engineered artifacts in current use.  Modern microprocessor chips
contain multiple processing units and multi-layer memories, for
instance, and they use complicated hardware/software strategies to
move data and threads of computation across those resources.  As a
result, the processor and memory loads during the execution of even a
very simple program can evolve in a very complicated fashion.
Figure~\ref{fig:ipc} shows a performance trace of a three-line loop
that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic \cite{mytkowicz09}, which places fundamental limits on
predictability.  Radical changes in this behavior can result if the
program is changed slightly.  The resulting traces span the whole
range of the complexity spectrum, from completely predictable to
completely unstructured, making this an ideal testbed for this study.

@@

@@

The rest of the paper is organized as follows:
Section~\ref{sec:related} discusses previous results on [[what]] and
[[what]] and situates this work in that
context. Section~\ref{sec:methods} covers the experimental setup and
methods used to collect each time series. Section~\ref{sec:model}
describes the prediction models we implemented.  In
Section~\ref{sec:meaComplex}, we review weighted permutation entropy.
In Section~\ref{sec:results}, we estimate the complexity of each time
series and compare that complexity to the accuracy of predictions
produced by the methods of Section~\ref{sec:model} on that time
series.  In Section~\ref{sec:conc}, we discuss these results and their
implications and consider future areas of research.

[[Following belongs in the methods section:]] In addition, some
performance traces even exhibit interesting regime changes as a
program switches between varying subroutines.

[[Following belongs in the conclusion.  Putting it here would be
    appropriate in a CS paper, but not in a physics journal, and not
    when we've just made the point that this is just a testbed.]] 
% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU, e.g., processor efficiency and similar statistics. Prediction
of these quantities at some finite time in the future, even with
perfect accuracy, does not violate the Rice-Shapiro
theorem~\cite{hopcroft} which essentially states that any non-trivial
property of a program is uncomputable.

[[Following is a good start at an abstract. It doesn't need to be
    reiterated in the introduction.]]  This paper provides insight
into when, why, and how forecast strategies fail when they are applied
to complicated time series by empirically estimating their inherent
complexity. [[Liz, do you think here would be better for the sentence
    about we are not the first, or is above where I asked Ryan to
    insert it ok? I worry it would muddy the water of this synopsis
    paragraph.]] We focus here on the specific example of computer
performance but believe the results apply to a much broader class of
time series. We conjecture that the complexity of traces from these
systems---which results from the inherent dimension, non-linearity,
and non-stationarity, as well as from measurement issues like noise,
aggregation, and finite data length---is \emph{empirically
  quantifiable}.  To validate this conjecture we model 120 different
performance traces using a variety of forecasting models.  We then
compare the empirically estimated \emph{complexity} with the accuracy
of each forecast method.
